---
title: "R Notebook"
output: html_notebook
---

Data:

-   rec: total receptions to filter out players with small sample size

-   air_yards: distance the ball traveled in the air before reaching them

-   yards_per_target: total yards / targets

-   yards_after_catch: yards after the ball was caught

-   td_rate: touchdowns per target

-   outside_rec: binary variable if the receiver caught the ball outside the hashes

-   dist_from_sticks: where they were targeted relative to distance needed to continue drive

```{r}
#rosters
rosters <- nflreadr::load_rosters(c(2010:2020))

rosters %>% 
  write_rds("~/Desktop/nfl_sql/nfl_sql/data/rosters_2010_2020.rds")

#players
players <- nflreadr::load_player_stats(seasons = c(2010:2020), file_type = 'rds')

players %>% 
  write_rds("~/Desktop/nfl_sql/nfl_sql/data/players_2010_2020.rds")

#pbp
#pbp <- nflreadr::load_pbp(2010:2020, file_type = 'rds')

all_pbp %>% 
  write_rds("data/2000-2019_pbp_raw.rds")

pbp <- all_pbp %>%
  filter(
    between(season, 2010, 2020) & 
      season_type == 'REG' & 
      !is.na(epa) & 
      !is.na(receiver_id)
    )

pbp %>% 
  write_rds("~/Desktop/nfl_sql/nfl_sql/data/pbp_2010_2020.rds")
```


```{r}
library(tidyverse)
```

```{r}
pbp <- pbp %>% 
  left_join(rosters %>% 
              select(position, season, gsis_id), by = c('fantasy_id' = 'gsis_id', 'season'))
```

```{r}
wr <- pbp %>% 
  filter(position == 'WR') 

wr %>% 
  write_rds("~/Desktop/nfl_sql/nfl_sql/data/wr_2010_2020.rds")
```

```{r}
wr_stats_full <- wr %>% 
  mutate(
    outside_pass = ifelse(pass_location != 'middle', 1, 0),
    pass_air_yards = ifelse(is.na(air_yards), 0, air_yards),
    pass_air_yards = ifelse(ydstogo <= 10, pass_air_yards, NA
                            )) %>% 
  group_by(receiver_id) %>% 
  summarise(
    rec = sum(complete_pass),
    air_yards = mean(pass_air_yards, na.rm = TRUE),
    yards_per_target = mean(yards_gained, na.rm = TRUE),
    yards_after_catch = mean(yards_after_catch, na.rm = TRUE),
    td_rate = mean(pass_touchdown),
    outside_rec = mean(outside_pass, na.rm = TRUE),
    dist_from_sticks = mean(pass_air_yards - ydstogo, na.rm = TRUE
                            )) %>% 
  filter(rec > 10) %>% 
  left_join(pbp %>% 
              count(receiver_id, receiver, posteam) %>% 
              group_by(receiver_id) %>% 
              arrange(-n) %>% 
              mutate(rn = row_number()) %>% 
              filter(rn == 1) %>% 
              select(-n, -rn)) %>%
  relocate(receiver, .before = rec)

wr_stats_full %>% 
  write_rds("~/Desktop/nfl_sql/nfl_sql/data/wr_stats_full_2010_2020.rds")
```

```{r}
wr_stats <- wr_stats_full %>% 
  select(-starts_with('rec'), -posteam) %>% 
  scale()

wr_stats %>% 
  as_tibble()
```

PCA from scratch

Singular value decomposition is a matrix factorization method that generalizes eigendecomposition of a matrix.
-   $U$
-   $\sum$ is the diagonal matrix containing the eigenvalues
-   $V$ contains the eigen vectors

```{r}
U <- svd(wr_stats)$u
D <- svd(wr_stats)$d
V <- svd(wr_stats)$v

DS <- diag(1/svd(wr_stats)$d[1:2])
US <- as.matrix(U[,1:2])
VS <- as.matrix(V[,1:2])


final <- VS %*% DS %*% t(US)

pca_data <- final[1:2, ] %>% MASS::ginv()

# Principal components explained variance
sum(D[[1]], D[[2]])/sum(D)
```

Using SVD to get the principal components, we can use the largest 2 to explain the ~54% of the variance. This allows for a 2 dimensional scatter plot to analyze players.

```{r}
pca_data %>% 
  as_tibble() %>% 
  bind_cols(wr_stats_full %>% 
              select(starts_with("rec"), posteam
                     )) %>% 
  ggplot(aes(V1, V2)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = paste(receiver, posteam)), max.overlaps = 10) +
  theme_minimal()
```




